--- 
title: "Classification and Regression by Random Forest"
author: "Payam Emami"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
link-citations: yes
description: "This is a short tutorial on how random forest works"
output:
  bookdown::html_document2:
    includes:
      in_header: header.html
  bookdown::gitbook:
    includes:
      in_header: header.html
---

# Prerequisites

In order to run the code in this chapter, you will need to install a number of packages. The packages are listed below.  

The **bookdown** package can be installed from CRAN or Github:

```{r eval=TRUE,each=TRUE,message=FALSE,warning=FALSE,error=FALSE,include=FALSE}

# Install packages
install.packages("bookdown")
install.packages("formatR")
install.packages("heplots")
install.packages("ipred")
install.packages("rpart")
install.packages("randomForest")
install.packages("readr")
install.packages("ggplot2")
install.packages("cowplot")
install.packages("mltools")
install.packages("tree")
install.packages("rfUtilities")
install.packages("reshape")
install.packages("gridBase")
install.packages("DiagrammeR")





```

```{r htmlTemp3, echo=FALSE, eval=TRUE,message=FALSE,warning=FALSE,error=FALSE,include=FALSE}
codejs <- readr::read_lines("js/codefolding.js")
collapsejs <- readr::read_lines("js/collapse.js")
transitionjs <- readr::read_lines("js/transition.js")

htmlhead <- 
  paste('
<script>',
paste(transitionjs, collapse = "\n"),
'</script>
<script>',
paste(collapsejs, collapse = "\n"),
'</script>
<script>',
paste(codejs, collapse = "\n"),
'</script>
<style type="text/css">
.code-folding-btn { margin-bottom: 4px; }
.row { display: flex; }
.collapse { display: none; }
.in { display:block }
</style>
<script>
$(document).ready(function () {
  window.initializeCodeFolding("show" === "hide");
});
</script>
', sep = "\n")

readr::write_lines(htmlhead, path = "header.html")
```


We will also use a dataset in this chapter. The data set contains the expression measure of 4 clinical variables, age and gender for 206 samples. Samples are from people with Alzheimer's disease (AD), frontotemporal Dementia (FTD), two groups of mild cognitive impairment (MCI), and non-demented controls. We will both the full dataset and also limited one (only AD and controls). Here is small description of the data from the original paper: "This study includes CSF samples from 76 AD patients, 74 mild cognitive impairment (MCI) patients, 11 frontotemporal dementia (FTD) patients, and 45 non-dementia controls. The MCI patients were followed for 4–8 years at 6–12 months intervals and eventually diagnosed with AD (MCI/AD converters) (n=21) or remained at the MCI stage (stable MCI) (n=53)"

Khoonsari, Payam Emami et al. ‘Improved Differential Diagnosis of Alzheimer’s Disease by Integrating ELISA and Mass Spectrometry-Based Cerebrospinal Fluid Biomarkers’. 1 Jan. 2019 : 639 – 651.

```{r,message=FALSE,warning=FALSE,error=FALSE}
# read the raw data
data<-read.csv("data/data.csv",sep = ";",
               stringsAsFactors = F,check.names = F,
               colClasses = c("numeric","character","numeric","numeric","numeric","numeric","numeric","character"))
# remove the index row
data<-data[,-1]
data$group[data$group=="non-demented controls"]<-"control"
# limit to AD and controls
limited_data<-data[data$group%in%c("AD","control"),]


```


```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```


